A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-12-08 article/01_Introduction.Rmd
started to prepare repo for code and data

Hardings narrative is, even when taking into account it's enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology: Instead of directly querying quantitative data the typical process for the creation of large scale overviews is to excerpt the literature of all relevant local research traditions. The present article in contrast explores whether openly available bulk radiocarbon data could allow to reconstruct the trends within the distribution of burial rites. This could set the precedent for for a more reproducible and statistically sound observation of the main shifts. The quantitative comparison of macroregional trends also paves the way for an analysis of patterns of interaction intensity on a continental scale.

@@

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase well-being and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite (or because?) their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

Burial traditions are ritualised behaviour most of the times heavily linked to religious and spiritual beliefs. For this paper they are assumed to be stylistic traits as defined by @dunnell1978style.

Fraser Neimans Article Stylistic Variation in Evolutionary Perspective: Inferences from Decorative Diversity and Interassemblage Distance in Illinois Woodland Ceramic Assemblages explores the new technologies proposed by Dunnell. stochastic processes and develops a simple framework for the simulation of propagation of non-functional ideas. In the process, the essential phenomena drift and flow (at Neiman: intergroup transmission) from population genetics in archaeology and to the interpretation of the Cultural distances between Indian ceramic inventories is used. Neiman's contribution was well received by the experts. very intensively received. Carl can be called Lipos works on ceramics in the Mississippi underflow, the Cultural Transmission Simulation with classical archaeological seriation, Shennans Investigation of the technological development of linear-band ceramics Ceramics in Merzbachtal or Timothy Kohlers analysis of the diversity of forms and its Role as proxy to the prehistoric social fabric in the southwest of the USA, again with ceramics . Also for the preparation of the simulation of this work was his contribution is an essential inspiration (see chapter 4.3.1).

>>>>

Hardings narrative is, even when taking into account it's enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology: Instead of directly querying quantitative data the typical process for the creation of large scale overviews is to excerpt the literature of all relevant local research traditions. The present article in contrast explores whether openly available bulk radiocarbon data could allow to reconstruct the trends within the distribution of burial rites. This could set the precedent for for a more reproducible and statistically sound observation of the main shifts. The quantitative comparison of macroregional trends also paves the way for an analysis of patterns of interaction intensity on a continental scale. A core question is how cultural and spatial distance are related over the course of the European Bronze Age.

@@

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase well-being and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite (or because?) their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

@neimanStylisticVariationEvolutionary1995 takes a closer look at this processes and develops a simple framework for the simulation of non-functional idea propagation. Neiman identifies the essential phenomena *drift* (random dominance of equivalent traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore he implements a measure of cultural distance to apply his theoretical insights to real world data. The workflow suggested by Neiman can be partially applied to the burial rites data examined in this article and was an important methodological inspiration: As burial traditions are ritualised behaviour and most of the time heavily linked to religious and spiritual beliefs they can be assumed to be stylistic traits as defined by @dunnell1978style. Besides @neimanStylisticVariationEvolutionary1995, another important template for the comparison of cultural and spatial was provided by @porcicSimulatingCulturalTransmission2014.
